---
title: 'Data Wrangling'
author: Divya Krishnan
date: 'Wednesday, October 14, 2015'
output: pdf_document
---

##### Instructions: #####
<hr> <!-- A horizontal line -->

Before beginning this assignment, please ensure you have access to R and RStudio. 

1. Download the `dataWrangling.Rmd` file from Canvas. Open `dataWrangling.Rmd` in RStudio and supply your solutions to the assignment by editing `dataWrangling.Rmd`.  You will also want to download the `weather.txt` data file, containing a dataset capturing daily temperatures in Cuernavaca, Mexico during 2010. 

2. Replace the "Insert Your Name Here" text in the `author:` field with your own full name.

3. Be sure to include code chucks, figures and written explanations as necessary. Any collaborators must be listed on the top of your assignment. Any figures should be clearly labeled and appropriately referenced within the text. 

5. When you have completed the assignment and have **checked** that your code both runs in the Console and knits correctly when you click `Knit HTML`, rename the R Markdown file to `YourLastName_YourFirstName.Rmd`, and submit on Canvas.

```{r Setup, message=FALSE}
# Stardard libraries
library(plyr)
library(dplyr)
library(ggplot2)
library(reshape2)
library(babynames)
```

#### Data Cleaning ####

In this problem we will use the `weather.txt` data. Import the data in **R** and answer the following questions.

```{r}
# Importing the weather dataset
weather<-read.delim(
  file = "weather.txt",
  header = TRUE,
  check.names = F
)

```

##### (a) What are the variables in this dataset? Describe what each variable measures. 

```{r}
# Variables in the dataset
colnames(weather)
summary(weather)
#View(weather)

```

##### (b) Tidy up the weather data such that each observation forms a row and each variable forms a column. You might find the following functions helpful:

- `melt` 
- `mutate`
- `dcast`

```{r}
# Tidying weather data using melt function
weather.tidy <- melt(
  data = weather,
  id = c("id","year","month","element"),
  variable.name=c("day"),
  value.name = "temperature"
)
#View(weather.tidy)
# Removing d prefix for day column
weather.tidy$day<-sub("d","",weather.tidy$day)
#Removing NAs
weather.tidy<-weather.tidy[complete.cases(weather.tidy[,6:6]),]

head(weather.tidy)
```
